Verification of a TCP/IP stack

### 2021-05-21 SecDev21/paper.tex
fixing ADA and SPARK section

Existing research deals with formally verifying several protocols TCP/IP stack levels. For example, the work in \emph{miTLS} \cite{bhargavan2013implementing} formally verifies an SSL/TLS protocol implementation, and the work in \cite{Reiher_2020} uses a technology called RecordFlux to safely parse data segments. Given the reliance of the other software-level protocols on the TCP protocol, formalizing the its specification and verifying its implementation is critical. Achieving these two goals in practice is challening. This is mainly because TCP's specification, \cite{rfc793}, is written in English rather than any formal language, leaving many parts underspecified. Moreover, the TCP protocol is inherently concurrent, making it complex to specify and verify. Previous work focused on formalizing the protocol to either prove key properties \cite{smith1996formal} or fully verify the TCP protocol's specifications \cite{ridge2008rigorous} using the HOL proof assistant tool. While for the later work, the authors claim that it is possible from their formalization to extract an actual implementation in Haskell, this has not been achieved yet. At this point of writing, we are not aware of any formally verified TCP protocol implementation.

@@

Ada is a general-purpose procedural programming language. The design of the Ada
language puts great emphasis on the safety and correctness of the program. This
objective is realized by using a readable syntax that uses keywords instead of
symbols where reasonable. The type system is strong and strict and many
potential violations of type constraints can be detected statically by the
compiler. If not, a run-time check is inserted into the program, to guarantee
the detection of incorrect situations. Ada 2012 introduced contract based
programming to Ada. In particular, it is possible to attach pre- and
postconditions to subprograms\footnote{In Ada, a distinction is made between
  functions that return a value, and procedures, which do
  not. \emph{Subprogram} is the term that designates both.}.  These conditions
can be checked during the execution of the program, just like assertions.

SPARK~\cite{mccormick_chapin_2015} is the name of a platform that provides
formal verification for Ada. It uses the user-provided contracts and attempts
to prove that the runtime checks cannot fail and that postconditions are
established by the corresponding subprograms.  As formal verification for the
whole Ada language would be intractable, SPARK is also the name of the subset
of the Ada language that is supported by the SPARK tool called
GNATprove\footnote{\url{http://docs.adacore.com/spark2014-docs/html/ug/}}.
This subset contains almost all features of Ada, though sometimes in a
restricted form.  In particular, expressions should be free from side effects,
and aliasing is forbidden (no two variables should share the same memory
location or overlap in memory), including when using pointers thanks to the use
of an ownership policy~\cite{dross2020recursive}.  This restriction greatly
simplifies the memory model used in the SPARK tool: any program variables can
be reasoned about independently from other variables.

GNATprove performs two kinds of analysis modularly on individual subprograms:
information flow to detect reads of uninitialized data and violations of flow
contracts; deductive verification based on the Why3 platform to generate
verification conditions for SMT solvers via a weakest-precondition calculus to
detect possible runtime errors and violations of functional contracts.

>>>>

Existing research deals with formally verifying several protocols TCP/IP stack levels. For example, the work in \emph{miTLS} \cite{bhargavan2013implementing} formally verifies an SSL/TLS protocol implementation, and the work in \cite{Reiher_2020} uses a technology called RecordFlux to safely parse data segments. Given the reliance of the other software-level protocols on the TCP protocol, formalizing the its specification and verifying its implementation is critical. Achieving these two goals in practice is challenging. This is mainly because TCP's specification, \cite{rfc793}, is written in English rather than any formal language, leaving many parts underspecified. Moreover, the TCP protocol is inherently concurrent, making it complex to specify and verify. Previous work focused on formalizing the protocol to either prove key properties \cite{smith1996formal} or fully verify the TCP protocol's specifications \cite{ridge2008rigorous} using the HOL proof assistant tool. While for the later work, the authors claim that it is possible from their formalization to extract an actual implementation in Haskell, this has not been achieved yet. At this point of writing, we are not aware of any formally verified TCP protocol implementation.

@@

Ada is a general-purpose, procedural programming language that puts great emphasis on the safety and correctness of the program. Some of the safety characteristics of the language are its strong typing system and its extensive compile-time and runtime checks. Common C programming vulnerabilities like using unsafe pointers or improper null termination of strings can not exist in an Ada program, while issues like buffer overflow or integer overflow can be dynamically captured by Ada’s runtime checks and dealt with via exception handlers. Furthermore, Ada introduced contracts, such as post- and pre-conditions, as part of the language’s standard syntax. Such contracts are vital for explicitly expressing software and verification requirements in the source code to allow for static analyzers or runtime checks to verify that the stated requirements are met.

By subsetting Ada, the SPARK programming language \cite{mccormick_chapin_2015} emerged to provide the largest possible subset of Ada, which is suitable for functional specification and static verification. For instance, in SPARK, expressions should be free from side effects, and aliasing is forbidden (no two variables should share the exact memory location or overlap in memory), including when using pointers thanks to the use of an ownership policy~\cite{dross2020recursive}.

\emph{GNATprove} \cite{GNATProve:users_manual} is the platform that provides formal verification for SPARK through two kinds of analysis. The first enables flow analysis to check the initialization of variables, look at data dependencies between inputs and outputs of subprograms, and detects unused assignments and unmodified variables. The second uses deductive verification to generate verification conditions for SMT solvers via a weakest-precondition calculus to detect possible runtime errors and violations of functional contracts.
